---
title: "Pavia 2018"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

## Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE, 
  dev = "png",
  dpi = 150,
  fig.align = "center",
  comment = NA
)
library(rstan)
library(dplyr)
library(lubridate)
library(ggplot2)
library(bayesplot)

theme_set(bayesplot::theme_default())

# seed for R's pseudo-RNGs, not Stan's
set.seed(1123) 
```

## The problem

### Background

Imagine that you are a statistician or data scientist working as an independent
contractor. One of your clients is a company that owns many residential buildings 
throughout New York City. The property manager explains that they are concerned about the number
of cockroach complaints that they receive from their buildings. Previously
the company has offered monthly visits from a pest inspector as a solution to
this problem. While this is the default solution of many property managers in
NYC, the tenants are rarely home when the inspector visits, and so the manager
reasons that this is a relatively expensive solution that is currently not very
effective.

One alternative to this problem is to deploy long term bait stations. In this
alternative, child and pet safe bait stations are installed throughout the
apartment building. Cockroaches obtain quick acting poison from these stations
and distribute it throughout the colony. The manufacturer of these bait stations
provides some indication of the space-to-bait efficacy, but the manager suspects
that this guidance was not calculated with NYC roaches in mind. NYC roaches, the
manager rationalizes, have more hustle than traditional roaches; and NYC
buildings are built differently than other common residential buildings in the
US. This is particularly important as the unit cost for each bait station per
year is quite high.

### The goal

The manager wishes to employ your services to help them to find the optimal
number of roach bait stations they should place in each of their buildings in
order to minimize the number of cockroach complaints while also keeping
expenditure on pest control affordable. 

A subset of the company's buildings have been randomly selected for an experiment: 

* At the beginning of each month, a pest inspector randomly places a number of
bait stations throughout the building, without knowledge of the current
cockroach levels in the building
* At the end of the month, the manager records
the total number of cockroach complaints in that building. 
* The manager would like to determine the optimal number of traps ($\textrm{traps}$) that
balances the lost revenue ($R$) that complaints ($\textrm{complaints}$) generate
with the all-in cost of maintaining the traps ($\textrm{TC}$). 

Fortunately, Bayesian data analysis provides a coherent framework for us to tackle this problem.

Formally, we are interested in finding

$$
\arg\max_{\textrm{traps} \in \mathbb{N}} \mathbb{E}_{\text{complaints}}[R(\textrm{complaints}(\textrm{traps})) - \textrm{TC}(\textrm{traps})]
$$

The property manager would also, if possible, like to learn how these results 
generalize to buildings they haven't treated so they can understand the
potential costs of pest control at buildings they are acquiring as well as for
the rest of their building portfolio.

As the property manager has complete control over the number of traps set, the
random variable contributing to this expectation is the number of complaints
given the number of traps. We will model the number of complaints as a function
of the number of traps.

## The data

The data provided to us is in a file called `pest_data.RDS`. Let's
load the data and see what the structure is:

```{r load-data}
pest_data <- readRDS('data/pest_data.RDS')
str(pest_data)
```

We have access to the following fields: 

* `complaints`: Number of complaints per building per month
* `building_id`: The unique building identifier
* `traps`: The number of traps used per month per building
* `date`: The date at which the number of complaints are recorded
* `live_in_super`: An indicator for whether the building as a live-in super
* `age_of_building`: The age of the building
* `total_sq_foot`: The total square footage of the building
* `average_tenant_age`: The average age of the tenants per building
* `monthly_average_rent`: The average monthly rent per building
* `floors`: The number of floors per building

First, let's see how many buildings we have data for:

```{r describe-data}
N_buildings <- length(unique(pest_data$building_id))
N_buildings
```

And make some plots of the raw data: 

```{r data-plots}
ggplot(pest_data, aes(x = complaints)) + 
  geom_bar()

ggplot(pest_data, aes(x = traps, y = complaints, color = live_in_super == TRUE)) + 
  geom_jitter()
```

```{r, data-plots-ts, fig.width = 6, fig.height = 8}
ggplot(pest_data, aes(x = date, y = complaints, color = live_in_super == TRUE)) + 
  geom_line(aes(linetype = "Number of complaints")) + 
  geom_point(color = "black") + 
  geom_line(aes(y = traps, linetype = "Number of traps"), color = "black", size = 0.25) + 
  facet_wrap(~building_id, scales = "free", ncol = 2, labeller = label_both) + 
  scale_x_date(name = "Month", date_labels = "%b") + 
  scale_y_continuous(name = "", limits = range(pest_data$complaints)) + 
  scale_linetype_discrete(name = "") + 
  scale_color_discrete(name = "Live-in super")
```


The first question we'll look at is just whether the number of complaints per
building per month is associated with the number of bait stations per building
per month, ignoring the temporal and across-building variation (we'll come back
to those sources of variation later in the document). That requires only two
variables, $\textrm{complaints}$ and $\textrm{traps}$. How should we model the
number of complaints?


## Bayesian workflow

See slides

## Modeling count data : Poisson distribution

We already know some rudimentary information about what we should expect. The
number of complaints over a month should be either zero or an integer. The
property manager tells us that it is possible but unlikely that number of
complaints in a given month is zero. Occasionally there are a very large number
of complaints in a single month. A common way of modeling this sort of skewed,
single bounded count data is as a Poisson random variable. One concern about
modeling the outcome variable as Poisson is that the data may be
over-dispersed, but we'll start with the Poisson model and then check 
whether over-dispersion is a problem by comparing our model's predictions 
to the data.

### Model 

Given that we have chosen a Poisson regression, we define the likelihood to be
the Poisson probability mass function over the number bait stations placed in
the building, denoted below as `traps`. This model assumes that the mean and
variance of the outcome variable `complaints` (number of complaints) is the
same. We'll investigate whether this is a good assumption after we fit the 
model.

For building $b = 1,\dots,10$ at time (month) $t = 1,\dots,12$, we have

$$
\begin{align*}
\textrm{complaints}_{b,t} & \sim \textrm{Poisson}(\lambda_{b,t}) \\
\lambda_{b,t} & = \exp{(\eta_{b,t})} \\
\eta_{b,t} &= \alpha + \beta \, \textrm{traps}_{b,t}
\end{align*}
$$

Let's encode this probability model in a Stan program. 

### Writing our first Stan model

* Write `simple_poisson_regression.stan` together

### Making sure our code is right

However, before we fit the model to real data, we should check that our
model works well with simulated data. We'll simulate data according to the model
and then check that we can sufficiently recover the parameter values used
in the simulation.

* Write `simple_poisson_regression_dgp.stan` together

We can use the `stan()` function to compile and fit the model, but here we will 
do the compilation and fitting in two stages to demonstrate what is really 
happening under the hood. 

First we will compile the Stan program (`simple_poisson_regression_dgp.stan`) that
will generate the fake data.

```{r , cache=TRUE, results="hide", message=FALSE}
comp_dgp_simple <- stan_model('stan_programs/simple_poisson_regression_dgp.stan')
```

Printing this object shows the Stan program: 

```{r}
print(comp_dgp_simple)
```

Now we can simulate the data by calling the `sampling()` function.

```{r runpoissondgp}
fitted_model_dgp <- sampling(
  comp_dgp_simple,
  data = list(N = nrow(pest_data), mean_traps = mean(pest_data$traps)),
  chains = 1,
  iter = 1,
  algorithm = 'Fixed_param',
  seed = 123
  )

# see http://mc-stan.org/rstan/articles/stanfit_objects.html for various
# ways of extracting the contents of the stanfit object
samps_dgp <- rstan::extract(fitted_model_dgp)
str(samps_dgp)
```


### Fit the model to the fake data:

In order to pass the fake data to our Stan program using RStan, we need to
arrange the data into a named list. The names must match the names used 
in the `data` block of the Stan program.

```{r}
stan_dat_fake <- list(
  N = nrow(pest_data), 
  traps = samps_dgp$traps[1, ], 
  complaints = samps_dgp$complaints[1, ]
)
str(stan_dat_fake)
```

Now that we have the simulated data we fit the model to see if we can recover
the `alpha` and `beta` parameters used in the simulation.

```{r , cache=TRUE, results="hide", message=FALSE}
comp_model_P <- stan_model('stan_programs/simple_poisson_regression.stan')
fit_model_P <- sampling(comp_model_P, data = stan_dat_fake, seed = 123)

posterior_alpha_beta <- as.matrix(fit_model_P, pars = c('alpha','beta'))
head(posterior_alpha_beta)
```


### Assess parameter recovery

```{r}
true_alpha_beta <- c(samps_dgp$alpha, samps_dgp$beta)
mcmc_recover_hist(posterior_alpha_beta, true = true_alpha_beta)
```

We don't do a great job recovering the parameters here simply because we're
simulating so few observations that the posterior uncertainty remains rather
large, but it looks at least _plausible_ ($\alpha$ and $\beta$ are contained
within the histograms). If we did the simulation with many more observations the
parameters would be estimated much more precisely.

We should also check if the `y_rep` datasets (in-sample predictions) that we coded 
in the `generated quantities` block are similar to the `y` (complaints) values
we conditioned on when fitting the model. (The **bayesplot**
package [vignettes](http://mc-stan.org/bayesplot/articles/graphical-ppcs.html)
are a good resource on this topic.)

Here is a plot of the density estimate of the observed data compared to 
200 of the `y_rep` datasets: 
```{r}
y_rep <- as.matrix(fit_model_P, pars = "y_rep")
ppc_dens_overlay(y = stan_dat_fake$complaints, yrep = y_rep[1:200, ])
```
In the plot above we have the kernel density estimate of the observed data ($y$,
thicker curve) and 200 simulated data sets ($y_{rep}$, thin curves) from the
posterior predictive distribution. If the model fits the data well, as it does
here, there is little difference between the observed dataset and the simulated
datasets.

Another plot we can make for count data is a rootogram. This is a plot of the
expected counts (continuous line) vs the observed counts (blue histogram). We
can see the model fits well because the observed histogram matches the expected
counts relatively well.

```{r}
ppc_rootogram(stan_dat_fake$complaints, yrep = y_rep)
```


### Fit with real data 

To fit the model to the actual observed data we'll first create a list to pass
to Stan using the variables in the `pest_data` data frame:

```{r stan-data}
stan_dat_simple <- list(
  N = nrow(pest_data), 
  complaints = pest_data$complaints,
  traps = pest_data$traps
)
```

As we have already compiled the model, we can jump straight to sampling from it.

```{r fit_P_real_data, cache=TRUE}
fit_P_real_data <- sampling(comp_model_P, data = stan_dat_simple)
```

and printing the parameters. What do these tell us? 

```{r results_simple_P}
print(fit_P_real_data, pars = c('alpha','beta'))
```

We can also plot the posterior distributions: 

```{r hist_simple_P}
mcmc_hist(as.matrix(fit_P_real_data, pars = c('alpha','beta')))
```

As we expected, it appears the number of bait stations set in a building is
associated with the number of complaints about cockroaches that were made in the
following month. However, we still need to consider how well the model fits.


### Posterior predictive checking

```{r}
y_rep <- as.matrix(fit_P_real_data, pars = "y_rep")
```

```{r marginal_PPC}
ppc_dens_overlay(y = stan_dat_simple$complaints, y_rep[1:200,])
```

As opposed to when we fit the model to simulated data above, here the simulated
datasets is not as dispersed as the observed data and don't seem to capture the
rate of zeros in the observed data. The Poisson model may not be sufficient for
this data.

Let's explore this further by looking directly at the proportion of zeros in the
real data and predicted data.
```{r}
prop_zero <- function(x) mean(x == 0)
ppc_stat(y = stan_dat_simple$complaints, yrep = y_rep, stat = "prop_zero")
```
The plot above shows the observed proportion of zeros (thick vertical line) and
a histogram of the proportion of zeros in each of the simulated data sets. It is
clear that the model does not capture this feature of the data well at all.

This next plot is a plot of the standardised residuals of the observed vs predicted number of complaints. 

```{r}
mean_y_rep <- colMeans(y_rep)
std_resid <- (stan_dat_simple$complaints - mean_y_rep) / sqrt(mean_y_rep)
qplot(mean_y_rep, std_resid) + hline_at(2) + hline_at(-2)
```

As you can see here, it looks as though we have more positive residuals than negative,
which indicates that the model tends to underestimate the number of complaints
that will be received.

The rootogram is another useful plot to compare the observed vs expected number
of complaints. This is a plot of the expected counts (continuous line) vs the
observed counts (blue histogram):

```{r}
ppc_rootogram(stan_dat_simple$complaints, yrep = y_rep)
```

If the model was fitting well these would be relatively similar, however in this
figure we can see the number of complaints is underestimated if there are few
complaints, over-estimated for medium numbers of complaints, and underestimated
if there are a large number of complaints.

We can also view how the predicted number of complaints varies with the number
of traps. From this we can see that the model doesn't seem to fully capture the
data.

```{r}
ppc_intervals(
  y = stan_dat_simple$complaints, 
  yrep = y_rep,
  x = stan_dat_simple$traps
) + 
  labs(x = "Number of traps", y = "Number of complaints")
```

Specifically, the model doesn't capture the tails of the observed data very
well.


## Expanding the model: multiple predictors

Modeling the relationship between complaints and bait stations is the simplest 
model. We can expand the model, however, in a few ways that will be beneficial 
for our client. Moreover, the manager has told us that they expect there are a
number of other reasons that one building might have more roach complaints than
another.

### Interpretability

Currently, our model's mean parameter is a rate of complaints per 30 days, but
we're modeling a process that occurs over an area as well as over time. We have
the square footage of each building, so if we add that information into the
model, we can interpret our parameters as a rate of complaints per square foot
per 30 days.

$$
\begin{align*}
\textrm{complaints}_{b,t} & \sim \textrm{Poisson}(\textrm{sq_foot}_b\,\lambda_{b,t}) \\
\lambda_{b,t} & = \exp{(\eta_{b,t} )} \\
\eta_{b,t} &= \alpha + \beta \, \textrm{traps}_{b,t}
\end{align*}
$$

The term $\text{sq_foot}$ is called an exposure term. If we log the term, we can 
put it in $\eta_{b,t}$:

$$
\begin{align*}
\textrm{complaints}_{b,t} & \sim \textrm{Poisson}(\lambda_{b,t}) \\
\lambda_{b,t} & = \exp{(\eta_{b,t} )} \\
\eta_{b,t} &= \alpha + \beta \, \textrm{traps}_{b,t} + \textrm{log_sq_foot}_b
\end{align*}
$$

A quick test shows us that there appears to be a relationship between the square
footage of the building and the number of complaints received:

```{r}
ggplot(pest_data, aes(x = log(total_sq_foot), y = log1p(complaints))) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```

Using the property manager's intuition, we include two extra pieces of
information we know about the building - the (log of the) square floor space and
whether there is a live in super or not - into both the simulated and real data.
 
```{r}
stan_dat_simple$log_sq_foot <- log(pest_data$total_sq_foot/1e4)
stan_dat_simple$live_in_super <- pest_data$live_in_super
```


### Stan program for Poisson multiple regression

Now we need a new Stan model that uses multiple predictors.

* Write `multiple_poisson_regression.stan` together

### Simulate fake data with multiple predictors

```{r compmultPDGP, cache=TRUE, results="hide", message=FALSE}
comp_dgp_multiple <- stan_model('stan_programs/multiple_poisson_regression_dgp.stan')
```

```{r runpoissondgp2}
fitted_model_dgp <-
  sampling(
  comp_dgp_multiple,
  data = list(N = nrow(pest_data)),
  chains = 1,
  cores = 1,
  iter = 1,
  algorithm = 'Fixed_param',
  seed = 123
  )
samps_dgp <- rstan::extract(fitted_model_dgp)
```

Now pop that simulated data into a list ready for Stan. 

```{r}
stan_dat_fake <- list(
  N = nrow(pest_data), 
  log_sq_foot = samps_dgp$log_sq_foot[1, ],
  live_in_super = samps_dgp$live_in_super[1, ],
  traps = samps_dgp$traps[1, ], 
  complaints = samps_dgp$complaints[1, ]
)
```

And then compile and fit the model we wrote for the multiple regression.

```{r, cache=TRUE, message=FALSE, warning=FALSE}
comp_model_P_mult <- stan_model('stan_programs/multiple_poisson_regression.stan')
fit_model_P_mult <- sampling(comp_model_P_mult, data = stan_dat_fake, chains = 4, cores = 4)
```

Then compare these parameters to the true parameters:

```{r}
posterior_alpha_beta <- as.matrix(fit_model_P_mult, pars = c('alpha','beta','beta_super'))
true_alpha_beta <- c(samps_dgp$alpha,samps_dgp$beta,samps_dgp$beta_super)
mcmc_recover_hist(posterior_alpha_beta, true = true_alpha_beta)
```

We've recovered the parameters sufficiently well, so we've probably coded the
Stan program correctly and we're ready to fit the real data.

### Fit the real data

Now let's use the real data and explore the fit.

```{r fit_mult_P_real_dat}
fit_model_P_mult_real <- sampling(comp_model_P_mult, data = stan_dat_simple)
y_rep <- as.matrix(fit_model_P_mult_real, pars = "y_rep")
ppc_dens_overlay(stan_dat_simple$complaints, y_rep[1:200,])
```

This again looks like we haven't captured the smaller counts very well, nor
have we captured the larger counts.


```{r}
prop_zero <- function(x) mean(x == 0)
ppc_stat(y = stan_dat_simple$complaints, yrep = y_rep, stat = "prop_zero", binwidth = 0.01)
```

We're still severely underestimating the proportion of zeros in the data. Ideally
this vertical line would fall somewhere within the histogram.

We can also plot uncertainty intervals for the predicted complaints for different
numbers of traps.

```{r}
ppc_intervals(
  y = stan_dat_simple$complaints, 
  yrep = y_rep,
  x = stan_dat_simple$traps
) + 
  labs(x = "Number of traps", y = "Number of complaints")
```

We can see that we've increased the tails a bit more at the larger numbers of traps
but we still have some large observed numbers of complaints that the model
would consider extremely unlikely events. 
 
## Modeling count data with the Negative Binomial

When we considered modelling the data using a Poisson, we saw that the model
didn't appear to fit as well to the data as we would like. In particular the
model underpredicted low and high numbers of complaints, and overpredicted the
medium number of complaints. This is one indication of over-dispersion, where
the variance is larger than the mean. A Poisson model doesn't fit over-dispersed
count data very well because the same parameter $\lambda$, controls both the
expected counts and the variance of these counts. The natural alternative to
this is the negative binomial model:

$$
\begin{align*}
\text{complaints}_{b,t} & \sim \text{Neg-Binomial}(\lambda_{b,t}, \phi) \\
\lambda_{b,t} & = \exp{(\eta_{b,t})} \\
\eta_{b,t} &= \alpha + \beta \, {\rm traps}_{b,t} + \beta_{\rm super} \, {\rm super}_{b} + \text{log_sq_foot}_{b}
\end{align*}
$$

In Stan the negative binomial mass function we'll use is called 
$\texttt{neg_binomial_2_log}(\text{ints} \, y, \text{reals} \, \eta, \text{reals} \, \phi)$ 
in Stan. Like the `poisson_log` function, this negative binomial mass function
that is parameterized in terms of its log-mean, $\eta$, but it also has a
precision $\phi$ such that

$$
\mathbb{E}[y] \, = \lambda = \exp(\eta)
$$

$$
\text{Var}[y] = \lambda + \lambda^2/\phi = \exp(\eta) + \exp(\eta)^2 / \phi.
$$ 

As $\phi$ gets larger the term $\lambda^2 / \phi$ approaches zero and so 
the variance of the negative-binomial approaches $\lambda$, i.e., the
negative-binomial gets closer and closer to the Poisson.

### Stan program for negative-binomial regression

* Write `multiple_NB_regression.stan` together


### Fake data fit: Multiple NB regression

```{r , cache=TRUE, results="hide", message=FALSE}
comp_dgp_multiple_NB <- stan_model('stan_programs/multiple_NB_regression_dgp.stan')
```

We're going to generate one draw from the fake data model so we can use the data
to fit our model and compare the known values of the parameters to the posterior
density of the parameters.

```{r fake_data_dgp_NB}
fitted_model_dgp_NB <-
  sampling(
  comp_dgp_multiple_NB,
  data = list(N = nrow(pest_data)),
  chains = 1,
  cores = 1,
  iter = 1,
  algorithm = 'Fixed_param',
  seed = 123
  )
samps_dgp_NB <- rstan::extract(fitted_model_dgp_NB)
```

Create a dataset to feed into the Stan model.

```{r NB_fake_stan_dat}
stan_dat_fake_NB <- list(
  N = nrow(pest_data), 
  log_sq_foot = samps_dgp_NB$log_sq_foot[1, ],
  live_in_super = samps_dgp_NB$live_in_super[1, ],
  traps = samps_dgp_NB$traps[1, ], 
  complaints = samps_dgp_NB$complaints[1, ]
)
```

Compile the inferential model.

```{r , cache=TRUE, results="hide", message=FALSE}
comp_model_NB <- stan_model('stan_programs/multiple_NB_regression.stan')
```

Now we run our NB regression over the fake data and extract the samples to
examine posterior predictive checks and to check whether we've sufficiently
recovered our known parameters, $\text{alpha}$ $\texttt{beta}$, .

```{r runNBoverfake, message=FALSE, warning=FALSE}
fitted_model_NB <- sampling(comp_model_NB, data = stan_dat_fake_NB, 
                            chains = 4, cores = 4)
posterior_alpha_beta_NB <- 
  as.matrix(fitted_model_NB,
            pars = c('alpha',
                     'beta',
                     'beta_super',
                     'inv_phi')
  )
```

Construct the vector of true values from your simulated dataset and compare to
the recovered parameters.
```{r}
true_alpha_beta_NB <- 
  c(samps_dgp_NB$alpha,
    samps_dgp_NB$beta,
    samps_dgp_NB$beta_super,
    samps_dgp_NB$inv_phi
  )
mcmc_recover_hist(posterior_alpha_beta_NB, true = true_alpha_beta_NB)
```


### Fit to real data and check the fit

```{r runNB}
fitted_model_NB <- sampling(comp_model_NB, data = stan_dat_simple)
samps_NB <- rstan::extract(fitted_model_NB)
```

Let's look at our predictions vs. the data.

```{r ppc-full}
y_rep <- samps_NB$y_rep
ppc_dens_overlay(stan_dat_simple$complaints, y_rep[1:200,])
```

It appears that our model now captures both the number of small counts better
as well as the tails. 

Let's check if the negative binomial model does a better job capturing
the number of zeros: 

```{r}
ppc_stat(y = stan_dat_simple$complaints, yrep = y_rep, stat = "prop_zero")
```

These look OK, but let's look at the standardized residual plot.

```{r}
mean_inv_phi <- mean(samps_NB$inv_phi)
mean_y_rep <- colMeans(y_rep)
std_resid <- (stan_dat_simple$complaints - mean_y_rep) / sqrt(mean_y_rep + mean_y_rep^2*mean_inv_phi)
qplot(mean_y_rep, std_resid) + hline_at(2) + hline_at(-2)
```

Looks OK, but we still have some very large _standardized_ residuals. This might be
because we are currently ignoring that the data are clustered by buildings, and
that the probability of roach issue may vary substantially across buildings.

```{r}
ppc_rootogram(stan_dat_simple$complaints, yrep = y_rep)
```

The rootogram now looks much more plausible. We can tell this because now the
expected number of complaints matches much closer to the observed number of
complaints. However, we still have some larger counts that appear
to be outliers for the model.

Check predictions by number of traps: 

```{r}
ppc_intervals(
  y = stan_dat_simple$complaints, 
  yrep = y_rep,
  x = stan_dat_simple$traps
) + 
  labs(x = "Number of traps", y = "Number of complaints")
```

We haven't used the fact that the data are clustered by building yet. A posterior 
predictive check might elucidate whether it would be a good idea to add the building
information into the model.

```{r ppc-group_means}
ppc_stat_grouped(
  y = stan_dat_simple$complaints, 
  yrep = y_rep, 
  group = pest_data$building_id, 
  stat = 'mean',
  binwidth = 0.2
)
```

We're getting plausible predictions for most building means but some are
estimated better than others and some have larger uncertainties than we might
expect. If we explicitly model the variation across buildings we may be able to
get much better estimates.

